---
title: "Right-Sized Intelligence: Using Model Capacity and Training Stage to Choose Effective Large Language Models"
output: html_document
date: "2026-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(lmtest)
library(sandwich)
library(stargazer)
library(ggplot2)
library(here)
library(broom)
library(dplyr)
library(data.table)
```

## Data Prep

```{r}
hf_dataset <- read.csv(here("data/interim/hf_dataset_clean.csv"))

# log transform model size based on scale and descriptive analysis check (parameters in billions)
hf_dataset$log_size <- log10(hf_dataset$x_params_b)

# Create Dummy Variable for Type (training stage) using str_detect to avoid emoji encoding issues
hf_dataset <- hf_dataset %>% 
  mutate(
    training_stage = case_when(
      str_detect(type, "fine-tuned")  ~ "Fine-tuned", 
      str_detect(type, "merges")      ~ "Base-Merge",
      str_detect(type, "chat")        ~ "Chat-Instruct",
      type == "🟢 pretrained"  ~ "Pretrained",
      TRUE ~ "Other"
    ),
    # Convert to factor so the linear model treats it as categorical
    training_stage = as.factor(training_stage)
  )
# Consistent colors for plots
stage_colors <- c(
  "Pretrained"     = "#E69F00",
  "Base-Merge"     = "#4A90D9",
  "Chat-Instruct"  = "#2C5F8A",
  "Fine-tuned"     = "#1A3A57"
)

hf_dataset %>% 
  group_by(training_stage) %>% 
  summarise(n = n())
```
```{r}
write.csv(hf_dataset, here("data/processed/hf_dataset.csv"), row.names = FALSE)
```

## Simple Model

```{r}
# Create simple baseline performance model of just model size. 
simple <- lm(mmlu_pro ~ log_size, data = hf_dataset)
summary(simple)

# Robust standard errors
simple_rse <- coeftest(simple, vcov = vcovHC(simple))
simple_rse
```
### Interpretation
10x increase in size = ~19.5 point increase. Not really trusting p-values, but this is in line directionally with scaling laws as understood

### Residuals vs Fitted
```{r}
# Plot size x by performance y with model line
ggplot(hf_dataset, aes(x = log_size, y = mmlu_pro)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1.2) +
  labs(title = "Simple Model: Performance by Model Size", 
       x = "Log10(Parameters)", 
       y = "MMLU-PRO Score") +
  theme_minimal()

# Fitted vs Residuals
hf_dataset_resid <- hf_dataset %>%
  mutate(
    simple_predictions = predict(simple),
    simple_residuals   = resid(simple)
  )

ggplot(hf_dataset_resid,aes(x = simple_predictions, y = simple_residuals)) + 
  geom_point() + geom_smooth(method = "loess") + theme_minimal()
```
Understimates anything smaller than ~300M parameters then a mix of pos/neg residuals


## Main Effect Model (Model 2)

```{r}
# Add training stage to model
model_2 <- lm(mmlu_pro ~ log_size + training_stage, data = hf_dataset)
summary(model_2)
# Robust SE
m2_rse <- coeftest(model_2, vcov = vcovHC(model_2))
m2_rse
```
### Interpretation

At 1B parameters (0 = 10^0 * Billion parameters): Merged models score 13 pts at 1B parameters, chat models score 10 points, fine-tuned 10, pretrained 6. A 10x size increase adds 18 points in base merges on average.

```{r}
# Scatterplot with interaction lines
df_no_other <- hf_dataset %>%
  filter(hf_dataset$training_stage != "Other")
ggplot(df_no_other, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(title = "Scaling Laws: Performance by Model Size and Training Stage",
       x = "Log10(Parameters)",
       y = "MMLU-PRO Score",
       color = "Training Stage") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  theme_minimal()

# Separate scatterplot with interaction lines by training stage
ggplot(df_no_other, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  facet_wrap(~training_stage) +
  labs(title = "Scaling Laws: Performance by Model Size and Training Stage",
       x = "Log10(Parameters)",
       y = "MMLU-PRO Score",
       color = "Training Stage") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  theme_minimal()
```
### F Test to compare iterations

```{r}
# F test comparing to the simple model using robust standard errors
waldtest(simple, model_2, vcov = vcovHC)
```
Model 2 has a better fit and is preferred


### Fitted vs Residuals
```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_2_predictions = predict(model_2),
    model_2_residuals   = resid(model_2)
  )


ggplot(hf_dataset_resid,aes(x = model_2_predictions, y = model_2_residuals)) + 
  geom_point() + geom_smooth(method = "loess") + theme_minimal()

```
Relatively consistent residuals from the last model. Still unreliable below predicted value of 0, closer to 0 when positive but with some variation. 

## Main Effects + Control (Model 3)

```{r}
# Add IFEval
model_3 <- lm(mmlu_pro ~ log_size + training_stage + if_eval, data = hf_dataset)
summary(model_3)
# Robust SE
m3_rse <- coeftest(model_3, vcov = vcovHC(model_3))
m3_rse
```
### Interpretation
At 1B parameters (0 = 10^0 * Billion parameters) and an IFEval score of 0: Merged models score 2 pts, chat models score -2 points, fine-tuned ~0, pretrained ~0. A 10x size increase adds 14 points in base merges on average. A 1 point increase in instruction following score equates to a small 0.3 increase in performance score (a 10pt increase = 3 points in performance, +30 IFEval = 10 MMLU-Pro).

### F Test to compare iterations
```{r}
# F test comparing to Model 2
waldtest(model_2, model_3, vcov = vcovHC)
```
Model 3 has a better fit and is preferred. 

### Residuals vs Fitted

```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_3_predictions = predict(model_3),
    model_3_residuals   = resid(model_3)
  )

ggplot(hf_dataset_resid,aes(x = model_3_predictions, y = model_3_residuals)) + 
  geom_point() + geom_smooth(method = "loess") + theme_minimal()
```
Positive prediction residual line getting flatter around zero, benchmark ranges now visible in upper and lower bands of residuals. 


## Interaction Model (Model 4)

```{r}
# Add the interaction of size and training stage. 
# Using '*' automatically includes log_size + training_stage + the interaction
model_4 <- lm(mmlu_pro ~ log_size * training_stage, data = hf_dataset)
summary(model_4)
# Robust SE
m4_rse <- coeftest(model_4, vcov = vcovHC(model_4))
m4_rse
```
### Interpretation

Interaction: fits a line for each training stage (i.e. base-merge is intercept + B1log_size, chat is (intercept + B2chat + (B1log_size + B6sizexchat)*size)) 
At 1B parameters (0 = 10^1 * billion parameters): base merges score 10 points, chat ~8, fine-tuned 9, pretrained ~8. 10x increase in parameters = 21 point increase in base merges, ~20 for chat, ~19 for fine-tuned, and 10 for pretrained. Major increase in training stage p-values. Only significant interaction (if we're trusting p-values at all) is the pretrained penalty for scaling. 

### F Test to compare model iterations (This model to model 2 bc IFEval model 3 isn't nested)

```{r}
# F test to see if interaction adds anything compared to just main effects
waldtest(model_2, model_4, vcov = vcovHC)
```
Model 4 has a better fit and is preferred. 

### Residuals vs Fitted

```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_4_predictions = predict(model_4),
    model_4_residuals   = resid(model_4)
  )

ggplot(hf_dataset_resid,aes(x = model_4_predictions, y = model_4_residuals)) + 
  geom_point() + geom_smooth(method = "loess") + theme_minimal()
```
Some variation away from zero added back in positive predictions.


## Full Control Model

```{r}
# Add the if_eval control variable to the interaction model
full <- lm(mmlu_pro ~ log_size * training_stage + if_eval, data = hf_dataset)
full_1 <- lm(mmlu_pro ~ log_size * training_stage + if_eval + I(log_size^2), data = hf_dataset)
summary(full_1)
# Robust SE
full_rse <- coeftest(full, vcov = vcovHC(full))
full_rse
```
### Intrpretation

At 1B parameters (0 = 10^1 * billion parameters) and 0 IFEval score: base merges score -1 points, chat ~3, fine-tuned 0, pretrained 2. 10x increase in parameters = 17 point increase in base merges, ~15 for chat, ~14 for fine-tuned, and 9 for pretrained. 1 pt increase in IFEval score equates to 0.3 pt increase in MMLU-Pro (or +10 points of IFEval = +3 points of performance, +30 IFEval = +10 MMLU-Pro). If we're trusting p-values at all, size and if_eval have extremely low p-values. Only significant interactions are the pretrained and to some degree fine-tuned penalties for scaling. Ignoring other since it's a very small amount of models (multi modal and continuously pretrained)

### F Test to compare iterations (full vs 3, full vs 4, full vs simple)

```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(model_3, full, vcov = vcovHC)
```
```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(model_4, full, vcov = vcovHC)
```

```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(simple, full, vcov = vcovHC)
```
Full model is a better fit and preferred over the simple model, model 3 and model 4.

### F Test Table

```{r}

w1 <- waldtest(simple,model_2,vcov = vcovHC)
w2 <- waldtest(model_2,model_3,vcov = vcovHC)
w3 <- waldtest(model_2,model_4,vcov = vcovHC)
w4 <- waldtest(model_3,full,vcov = vcovHC)
w5 <- waldtest(model_4,full,vcov = vcovHC)
w6 <- waldtest(simple,full,vcov = vcovHC)

F_comparison_table <- rbind(
  tidy(w1)[2, ] %>% mutate(Comparison = "simple vs model_2"),
  tidy(w2)[2, ] %>% mutate(Comparison = "model_2 vs model_3"),
  tidy(w3)[2, ] %>% mutate(Comparison = "model_2 vs model_4"),
  tidy(w4)[2, ] %>% mutate(Comparison = "model_3 vs full"),
  tidy(w5)[2, ] %>% mutate(Comparison = "model_4 vs full"),
  tidy(w6)[2, ] %>% mutate(Comparison = "simple vs full"),
deparse.level = 1) %>%
  select(Comparison, statistic, p.value, df, df.residual)

print(F_comparison_table)

# Prep Wald Results for Stargazer Regression table
f1 <- round(w1$F[2],3)
p1 <- w1$`Pr(>F)`[2]
f2 <- round(w2$F[2],3)
p2 <- w2$`Pr(>F)`[2]
f3 <- round(w3$F[2],3)
p3 <- w3$`Pr(>F)`[2]
f4 <- round(w4$F[2],3)
p4 <- w4$`Pr(>F)`[2]
f5 <- round(w5$F[2],3)
p5 <- w5$`Pr(>F)`[2]
f6 <- round(w6$F[2],3)
p6 <- w6$`Pr(>F)`[2]

get_stars <- function(p) {
  if (p < 0.01) return("***")
  if (p < 0.05) return("**")
  if (p < 0.1)  return("*")
  return("")
}

wald_row <- c("Wald F-test",
              " ",
              paste0(f1,get_stars(p1)), 
              paste0(f2,get_stars(p2)),
              paste0(f3,get_stars(p3)),
              #paste0(f4,get_stars(p4)),
              paste0(f5,get_stars(p5))#,
              #paste0(f6,get_stars(p6))
              )

```
### Residuals vs Fitted

```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    full_predictions = predict(full),
    full_residuals   = resid(full)
  )

ggplot(hf_dataset_resid,aes(x = full_predictions, y = full_residuals)) + 
  geom_point() + geom_smooth(method = "loess") +
  labs(title = "Residuals vs Fitted Values", y="Residuals", x= "Fitted Values") +
  theme_minimal()

full_preds <- predict(full, new.data=hf_dataset, model_type="prediction",interval="confidence", level = 0.95)
hf_dataset_preds <- cbind(hf_dataset,full_preds)
hf_dataset_preds <- hf_dataset_preds %>%
  filter(hf_dataset_preds$training_stage != "Other")
```
Average residuals hover around 0 for positive predicted values. Range of performance benchmarks visible in top and bottom band.

## Predicted Values Across Size and Training Stage

```{r}
# expand.grid method and data.table method
# predicted values across size, if_eval held at median
pred_grid <- expand.grid(
  log_size = seq(min(hf_dataset$log_size),
                 max(hf_dataset$log_size),
                 length.out = 100),
  training_stage = levels(hf_dataset$training_stage),
  if_eval = 45.66 # median IFEval score
) %>%
  filter(training_stage != "Other")

pred_grid$pred_mmlu_pro <- predict(full, newdata = pred_grid)

pred_ci <- predict(full, newdata = pred_grid, interval = "confidence", level = 0.95)
pred_grid <- cbind(pred_grid, pred_ci)

  # x-axis labels 
  x_breaks <- c(seq(0, 0.8, 0.2), 0.95)
  x_labels <- c("0", "20", "40", "60", "80", "100%")

  # y-axis labels 
  y_breaks <- seq(0.2, 0.8, by = 0.1)
  y_labels <- c("20", "30", "40", "50", "60", "70", "80%")

# creating predictions and CI
pred_data <- data.frame( 
  training_stage = levels(hf_dataset$training_stage), 
  log_size = rep(seq(from = min(hf_dataset$log_size, na.rm = T), to = max(hf_dataset$log_size, na.rm = T), length.out = 100) , 5),
  if_eval = 45.66 # median IFEval score
)

pred_data$fit     <- predict.lm(full, newdata = pred_data, interval = 'confidence')[ , 'fit']
pred_data$se.fit  <- predict.lm(full, newdata = pred_data, se.fit = T,)$se.fit
pred_data$low.ci  <- predict.lm(full, newdata = pred_data, interval = 'confidence', level = 0.95)[ , 'lwr']
pred_data$high.ci <- predict.lm(full, newdata = pred_data, interval = 'confidence', level = 0.95)[ , 'upr']

pred_data <- data.table(pred_data) %>%
  filter(training_stage != "Other")

ggplot(pred_data, aes(x = log_size, y = fit, color = training_stage, fill = training_stage)) +
 # geom_line(data = dens_df, aes(x = x, y = y), linewidth = 0.7, color = "black") +
  geom_ribbon(aes(x=log_size,ymin = low.ci, ymax = high.ci), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 25.84, linetype = "dashed", color = "grey50") +
  annotate("text", x=-2, y=29, label="Mean MMLU-Pro Score", color="grey50") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  scale_x_continuous(
    breaks = c(seq(-2,1,1),2),
    labels = c("10M","100M","1B","10B","100B")
  ) +
  labs(
    title = "Predicted Performance By Training Stage",
    subtitle = "Full Model Prediction across LLM Size",
    caption = "Source: HuggingFace Open LLM Leaderboard Contents Dataset 2024",
    x = "Parameter Count (log)",
    y = "Predicted MMLU-Pro",
    color = "Training Stage",
    fill = "Training Stage"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"), plot.caption = element_text(hjust = 0))
ggplot(pred_grid, aes(x = log_size, y = fit, color = training_stage, fill = training_stage)) +
#  geom_line(data = dens_df, aes(x = x, y = y), linewidth = 0.7, color = "black") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 25.84, linetype = "dashed", color = "grey50") +
  annotate("text", x=-2, y=29, label="Mean MMLU-Pro Score", color="grey50") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  scale_x_continuous(
    breaks = c(seq(-2,1,1),2),
    labels = c("10M","100M","1B","10B","100B")
  ) +
  labs(
    title = "Predicted Performance By Training Stage",
    subtitle = "Full Model Prediction across LLM Size",
    caption = "Source: HuggingFace Open LLM Leaderboard Contents Dataset 2024",
    x = "Parameter Count (log)",
    y = "Predicted MMLU-Pro",
    color = "Training Stage",
    fill = "Training Stage"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"), plot.caption = element_text(hjust = 0))
```
Size is the biggest impact on performance, but has diminishing returns with the level-log relationship. So anything 5B parameters or below, you might as well go with pretrained and your performance won't be amazing, but it will be ok. 5-10B you can see meaningful performance gains (average to better than average performance) and the closer you are to 10B the more it makes sense for you to go with a merged model (or even chat/fine tune). To get the same performance gain as you got from 1B to 10B, you'd have to get to 100B. the higher you go, the more it makes sense to go with a merged model (or chat/fine tuned). So I think maybe the "sweet spot" we're looking for could be around 10B merged model, still on the small side and above average performance

### Fitted vs Residuals by Training Stage

```{r}
ggplot(hf_dataset_resid, aes(x = full_predictions, y = full_residuals)) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.2) + 
  facet_wrap(~training_stage) + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted by Training Stage",
       x = "Fitted Values",
       y = "Residuals")

```
Hover around zero with more variation for pretrained (smaller group)

##Q-Q Plot
```{r}
hist(full_rse)
qqnorm(full_rse)
qqline(full_rse, lwd=2)
```
The distribution has "fat tails" meaning more variance at the low and high end. 

## Descriptive Statistics Table
```{r results='asis', echo=FALSE}
desc_stats <- hf_dataset %>% 
  group_by(training_stage) %>% 
  summarise(n = n(), 
            avg_size_B = mean(x_params_b),
            size_range = paste0(min(x_params_b), " - ", max(x_params_b)),
            avg_mmlu_pro = mean(mmlu_pro),
            avg_if_eval = mean(if_eval)
  )

desc_stats
```


## Regression Table Output

```{r results='asis', echo=FALSE}
# Extract the robust standard errors column from each coeftest object
se_simple <- simple_rse[, 2]
se_m2 <- m2_rse[, 2]
se_m3 <- m3_rse[, 2]
se_m4 <- m4_rse[, 2]
se_full <- full_rse[, 2]

suppressWarnings(
  stargazer(simple, model_2, model_3, model_4, full, 
            type = "html", 
            title = "Model Performance Across Size, Training Stage, and Instruction Following",
            dep.var.labels = "MMLU-PRO Score",
            column.labels = c("Simple", "Main Effects", "+IFEval", "Interaction", "Full Model"),
            covariate.labels = c("Log(Parameters)", 
                                 "Chat-Instruct", 
                                 "Fine-tuned", 
                                 "Other", 
                                 "Pretrained", 
                                 "IFEval Score", 
                                 "Log(Params) × Chat-Instruct", 
                                 "Log(Params) × Fine-tuned", 
                                 "Log(Params) × Other", 
                                 "Log(Params) × Pretrained", 
                                 "Constant"),

            omit.stat = "f",
            add.lines = list(wald_row),
            notes = "Note: 'Base-Merge' is the reference category for Training Stage. Robust standard errors (HC3) reported in parentheses.",
            se = list(se_simple, se_m2, se_m3, se_m4, se_full), 
            header = FALSE)
)
