---
title: "Right-Sized Intelligence: Using Model Capacity and Training Stage to Choose Effective Large Language Models"
output: html_document
date: "2026-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(lmtest)
library(sandwich)
library(stargazer)
library(ggplot2)
library(here)
```

## Data Prep

```{r}
hf_dataset <- read.csv(here("data/interim/hf_dataset_clean.csv"))

# log transform model size based on scale and descriptive analysis check (parameters in billions)
hf_dataset$log_size <- log10(hf_dataset$x_params_b)

# Create Dummy Variable for Type (training stage) using str_detect to avoid emoji encoding issues
hf_dataset <- hf_dataset %>% 
  mutate(
    training_stage = case_when(
      str_detect(type, "fine-tuned") ~ "Fine-tuned", 
      str_detect(type, "merges")     ~ "Base-Merge",
      str_detect(type, "chat")       ~ "Chat-Instruct",
      str_detect(type, "pretrained") ~ "Pretrained",
      TRUE ~ "Other"
    ),
    # Convert to factor so the linear model treats it as categorical
    training_stage = as.factor(training_stage)
  )
```

## Simple Model

```{r}
# Create simple baseline performance model of just model size. 
simple <- lm(mmlu_pro ~ log_size, data = hf_dataset)
summary(simple)

simple_rse <- coeftest(simple, vcov = vcovHC(simple))
simple_rse
```
10x increase in size = ~19.5 point increase. Not really trusting p-values, but this is in line directionally with scaling laws as understood

```{r}
# Plot size x by performance y with model line
ggplot(hf_dataset, aes(x = log_size, y = mmlu_pro)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1.2) +
  labs(title = "Simple Model: Performance by Model Size", 
       x = "Log10(Parameters)", 
       y = "MMLU-PRO Score") +
  theme_minimal()

# Fitted vs Residuals
plot(simple, which = 1, main = "Simple Model: Residuals vs Fitted", pch=16, col="darkorange")
```
Understimates anything smaller than ~300M parameters then a mix of pos/neg residuals


## Main Effect Model (Model 2)

```{r}
model_2 <- lm(mmlu_pro ~ log_size + training_stage, data = hf_dataset)
summary(model_2)

m2_rse <- coeftest(model_2, vcov = vcovHC(model_2))
m2_rse
```
At 1B parameters (0 = 10^0 * Billion parameters): Merged models score 13 pts at 1B parameters, chat models score 10 points, fine-tuned 10, pretrained 6. A 10x size increase adds 18 points in base mergeson average.

```{r}
# Scatterplot with interaction lines
ggplot(hf_dataset, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(title = "Scaling Laws: Performance by Model Size and Training Stage",
       x = "Log10(Parameters)",
       y = "MMLU-PRO Score",
       color = "Training Stage") +
  theme_minimal()
```


```{r}
# F test comparing to the simple model using robust standard errors
waldtest(simple, model_2, vcov = vcovHC)
```
Model 2 has a better fit and is preferred

```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_2_predictions = predict(model_2),
    model_2_residuals   = resid(model_2)
  )

plot_model_2a <- hf_dataset_resid %>%
  ggplot(aes(x = log_size, y = model_2_residuals)) + 
  geom_point() + stat_smooth()

plot_model_2b <- hf_dataset_resid %>%
  ggplot(aes(x = model_2_predictions, y = model_2_residuals)) + 
  geom_point() + stat_smooth()

plot_model_2a
plot_model_2b

```


## Main Effects + Control (Model 3)

```{r}
model_3 <- lm(mmlu_pro ~ log_size + training_stage + if_eval, data = hf_dataset)
summary(model_3)

m3_rse <- coeftest(model_3, vcov = vcovHC(model_3))
m3_rse
```
At 1B parameters (0 = 10^0 * Billion parameters) and an IFEval score of 0: Merged models score 2 pts, chat models score -2 points, fine-tuned ~0, pretrained ~0. A 10x size increase adds 14 points in base merges on average. A 1 point increase in instruction following score equates to a small 0.3 increase in performance score (a 10pt increase = 3 points in performance, +30 IFEval = 10 MMLU-Pro).


```{r}
# F test comparing to Model 2
waldtest(model_2, model_3, vcov = vcovHC)
```
Model 3 has a better fit and is preferred. 

```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_3_predictions = predict(model_3),
    model_3_residuals   = resid(model_3)
  )
3

plot_model_3 <- hf_dataset_resid %>%
  ggplot(aes(x = model_3_predictions, y = model_3_residuals)) + 
  geom_point() + stat_smooth()

plot_model_3
```



## Interaction Model (Model 4)

```{r}
# Add the interaction of size and training stage. 
# Using '*' automatically includes log_size + training_stage + the interaction
model_4 <- lm(mmlu_pro ~ log_size * training_stage, data = hf_dataset)
summary(model_4)

m4_rse <- coeftest(model_4, vcov = vcovHC(model_4))
m4_rse
```
Interaction: fits a line for each training stage (i.e. base-merge is intercept + B1log_size, chat is (intercept + B2chat + (B1log_size + B6sizexchat)*size)) 
At 1B parameters (0 = 10^1 * billion parameters): base merges score 10 points, chat ~8, fine-tuned 9, pretrained ~8. 10x increase in parameters = 21 point increase in base merges, ~20 for chat, ~19 for fine-tuned, and 10 for pretrained. Major increase in training stage p-values. Only significant interaction (if we're trusting p-values at all) is the pretrained penalty for scaling. 


```{r}
# F test to see if interaction adds anything compared to just main effects
waldtest(model_2, model_4, vcov = vcovHC)
```
Model 4 has a better fit and is preferred. 
```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_4_predictions = predict(model_4),
    model_4_residuals   = resid(model_4)
  )
3

plot_model_4 <- hf_dataset_resid %>%
  ggplot(aes(x = model_4_predictions, y = model_4_residuals)) + 
  geom_point() + stat_smooth()

plot_model_4
```



## Full Control Model

```{r}
# Add the if_eval control variable to the interaction model
full <- lm(mmlu_pro ~ log_size * training_stage + if_eval, data = hf_dataset)
summary(full)

full_rse <- coeftest(full, vcov = vcovHC(full))
full_rse
```
At 1B parameters (0 = 10^1 * billion parameters) and 0 IFEval score: base merges score -1 points, chat ~3, fine-tuned 0, pretrained 2. 10x increase in parameters = 17 point increase in base merges, ~15 for chat, ~14 for fine-tuned, and 9 for pretrained. 1 pt increase in IFEval score equates to 0.3 pt increase in MMLU-Pro (or +10 points of IFEval = +3 points of performance, +30 IFEval = +10 MMLU-Pro). If we're trusting p-values at all, size and if_eval have extremely low p-values. Only significant interactions are the pretrained and to some degree fine-tuned penalties for scaling. Ignoring other since it's a very small amount of models (multi modal and continuously pretrained)


```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(model_3, full, vcov = vcovHC)
```
```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(model_4, full, vcov = vcovHC)
```

```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(simple, full, vcov = vcovHC)
```
Full model is a better fit and preferred over the simple model, model 3 and model 4.

```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    full_predictions = predict(full),
    full_residuals   = resid(full)
  )
3

plot_full <- hf_dataset_resid %>%
  ggplot(aes(x = full_predictions, y = full_residuals)) + 
  geom_point() + stat_smooth()

plot_full
```



```{r}
# Scatterplot with interaction lines
ggplot(hf_dataset, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(title = "Scaling Laws: Performance by Model Size and Training Stage",
       x = "Log10(Parameters)",
       y = "MMLU-PRO Score",
       color = "Training Stage") +
  theme_minimal()

# Fitted vs Residuals for Full Model
plot(full, which = 1, main = "Full Model: Residuals vs Fitted", pch=16, col="purple")
```

## Regression Table Output

```{r results='asis', echo=FALSE}
# Extract the robust standard errors column from each coeftest object
se_simple <- simple_rse[, 2]
se_m2 <- m2_rse[, 2]
se_m3 <- m3_rse[, 2]
se_m4 <- m4_rse[, 2]
se_full <- full_rse[, 2]

suppressWarnings(
  stargazer(simple, model_2, model_3, model_4, full, 
            type = "html", 
            title = "Model Performance Across Size, Training Stage, and Instruction Following",
            dep.var.labels = "MMLU-PRO Score",
            column.labels = c("Simple", "Main Effects", "+ IFEval", "Interaction", "Full Model"),
            covariate.labels = c("Log(Parameters)", 
                                 "Chat-Instruct", 
                                 "Fine-tuned", 
                                 "Other", 
                                 "Pretrained", 
                                 "IFEval Score", 
                                 "Log(Params) × Chat-Instruct", 
                                 "Log(Params) × Fine-tuned", 
                                 "Log(Params) × Other", 
                                 "Log(Params) × Pretrained", 
                                 "Base-Merge"),
            notes = "Note: 'Base-Merge' is the reference category for Training Stage. Robust standard errors (HC3) reported in parentheses.",
            se = list(se_simple, se_m2, se_m3, se_m4, se_full), 
            header = FALSE)
)
