---
title: "Right-Sized Intelligence: Using Model Capacity and Training Stage to Choose Effective Large Language Models"
output: html_document
date: "2026-04-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(lmtest)
library(sandwich)
library(stargazer)
library(ggplot2)
library(here)
```

## Data Prep

```{r}
hf_dataset <- read.csv(here("data/interim/hf_dataset_clean.csv"))

# log transform model size based on scale and descriptive analysis check (parameters in billions)
hf_dataset$log_size <- log10(hf_dataset$x_params_b)

# Create Dummy Variable for Type (training stage) using str_detect to avoid emoji encoding issues
hf_dataset <- hf_dataset %>% 
  mutate(
    training_stage = case_when(
      str_detect(type, "fine-tuned")  ~ "Fine-tuned", 
      str_detect(type, "merges")      ~ "Base-Merge",
      str_detect(type, "chat")        ~ "Chat-Instruct",
      type == "🟢 pretrained"  ~ "Pretrained",
      #str_detect(type, "continuously")~ "Other",
      #str_detect(type, "multimodal")  ~ "Other",
      TRUE ~ "Other"
    ),
    # Convert to factor so the linear model treats it as categorical
    training_stage = as.factor(training_stage)
  )
hf_dataset %>% 
  group_by(training_stage) %>% 
  summarise(n = n())
```

## Simple Model

```{r}
# Create simple baseline performance model of just model size. 
simple <- lm(mmlu_pro ~ log_size, data = hf_dataset)
summary(simple)

simple_rse <- coeftest(simple, vcov = vcovHC(simple))
simple_rse
```
10x increase in size = ~19.5 point increase. Not really trusting p-values, but this is in line directionally with scaling laws as understood

```{r}
# Plot size x by performance y with model line
ggplot(hf_dataset, aes(x = log_size, y = mmlu_pro)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 1.2) +
  labs(title = "Simple Model: Performance by Model Size", 
       x = "Log10(Parameters)", 
       y = "MMLU-PRO Score") +
  theme_minimal()

# Fitted vs Residuals
plot(simple, which = 1, main = "Simple Model: Residuals vs Fitted", pch=16, col="darkorange")
```
Understimates anything smaller than ~300M parameters then a mix of pos/neg residuals


## Main Effect Model (Model 2)

```{r}
model_2 <- lm(mmlu_pro ~ log_size + training_stage, data = hf_dataset)
summary(model_2)

m2_rse <- coeftest(model_2, vcov = vcovHC(model_2))
m2_rse
```
At 1B parameters (0 = 10^0 * Billion parameters): Merged models score 13 pts at 1B parameters, chat models score 10 points, fine-tuned 10, pretrained 6. A 10x size increase adds 18 points in base mergeson average.

```{r}
# Scatterplot with interaction lines
df_no_other <- hf_dataset %>%
  filter(hf_dataset$training_stage != "Other")
ggplot(df_no_other, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  #geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(title = "Scaling Laws: Performance by Model Size and Training Stage",
       x = "Log10(Parameters)",
       y = "MMLU-PRO Score",
       color = "Training Stage") +
  theme_minimal()
```


```{r}
# F test comparing to the simple model using robust standard errors
waldtest(simple, model_2, vcov = vcovHC)
```
Model 2 has a better fit and is preferred

```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_2_predictions = predict(model_2),
    model_2_residuals   = resid(model_2)
  )

plot_model_2a <- hf_dataset_resid %>%
  ggplot(aes(x = log_size, y = model_2_residuals)) + 
  geom_point() + stat_smooth()

plot_model_2b <- hf_dataset_resid %>%
  ggplot(aes(x = model_2_predictions, y = model_2_residuals)) + 
  geom_point() + stat_smooth()

plot_model_2a
plot_model_2b

```


## Main Effects + Control (Model 3)

```{r}
model_3 <- lm(mmlu_pro ~ log_size + training_stage + if_eval, data = hf_dataset)
summary(model_3)

m3_rse <- coeftest(model_3, vcov = vcovHC(model_3))
m3_rse
```
At 1B parameters (0 = 10^0 * Billion parameters) and an IFEval score of 0: Merged models score 2 pts, chat models score -2 points, fine-tuned ~0, pretrained ~0. A 10x size increase adds 14 points in base merges on average. A 1 point increase in instruction following score equates to a small 0.3 increase in performance score (a 10pt increase = 3 points in performance, +30 IFEval = 10 MMLU-Pro).


```{r}
# F test comparing to Model 2
waldtest(model_2, model_3, vcov = vcovHC)
```
Model 3 has a better fit and is preferred. 

```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_3_predictions = predict(model_3),
    model_3_residuals   = resid(model_3)
  )
3

plot_model_3 <- hf_dataset_resid %>%
  ggplot(aes(x = model_3_predictions, y = model_3_residuals)) + 
  geom_point() + stat_smooth()

plot_model_3
```



## Interaction Model (Model 4)

```{r}
# Add the interaction of size and training stage. 
# Using '*' automatically includes log_size + training_stage + the interaction
model_4 <- lm(mmlu_pro ~ log_size * training_stage, data = hf_dataset)
summary(model_4)

m4_rse <- coeftest(model_4, vcov = vcovHC(model_4))
m4_rse
```
Interaction: fits a line for each training stage (i.e. base-merge is intercept + B1log_size, chat is (intercept + B2chat + (B1log_size + B6sizexchat)*size)) 
At 1B parameters (0 = 10^1 * billion parameters): base merges score 10 points, chat ~8, fine-tuned 9, pretrained ~8. 10x increase in parameters = 21 point increase in base merges, ~20 for chat, ~19 for fine-tuned, and 10 for pretrained. Major increase in training stage p-values. Only significant interaction (if we're trusting p-values at all) is the pretrained penalty for scaling. 


```{r}
# F test to see if interaction adds anything compared to just main effects
waldtest(model_2, model_4, vcov = vcovHC)
```
Model 4 has a better fit and is preferred. 

```{r}
# F test to compare Ifeval vs interaction X
anova(model_4, model_3, test = "F" )
```



```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    model_4_predictions = predict(model_4),
    model_4_residuals   = resid(model_4)
  )

plot_model_4 <- hf_dataset_resid %>%
  ggplot(aes(x = model_4_predictions, y = model_4_residuals)) + 
  geom_point() + stat_smooth()

plot_model_4
```



## Full Control Model

```{r}
# Add the if_eval control variable to the interaction model
full <- lm(mmlu_pro ~ log_size * training_stage + if_eval, data = hf_dataset)
full_1 <- lm(mmlu_pro ~ log_size * training_stage + if_eval + I(log_size^2), data = hf_dataset)
summary(full_1)

full_rse <- coeftest(full, vcov = vcovHC(full))
full_rse
```
At 1B parameters (0 = 10^1 * billion parameters) and 0 IFEval score: base merges score -1 points, chat ~3, fine-tuned 0, pretrained 2. 10x increase in parameters = 17 point increase in base merges, ~15 for chat, ~14 for fine-tuned, and 9 for pretrained. 1 pt increase in IFEval score equates to 0.3 pt increase in MMLU-Pro (or +10 points of IFEval = +3 points of performance, +30 IFEval = +10 MMLU-Pro). If we're trusting p-values at all, size and if_eval have extremely low p-values. Only significant interactions are the pretrained and to some degree fine-tuned penalties for scaling. Ignoring other since it's a very small amount of models (multi modal and continuously pretrained)


```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(model_3, full, vcov = vcovHC)
```
```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(model_4, full, vcov = vcovHC)
```

```{r}
# Final F-test: Does the full model perform significantly better than the simple baseline?
waldtest(simple, full, vcov = vcovHC)
```
Full model is a better fit and preferred over the simple model, model 3 and model 4.


```{r}
results <- waldtest(simple, model_2, model_3, full,vcov = vcovHC)
print(results)

results_4 <- waldtest(simple, model_2, model_4, full,vcov = vcovHC)
print(results_4)

library(broom)
library(dplyr)

w1 <- waldtest(simple,model_2,vcov = vcovHC)
w2 <- waldtest(model_2,model_3,vcov = vcovHC)
w3 <- waldtest(model_2,model_4,vcov = vcovHC)
w4 <- waldtest(model_3,full,vcov = vcovHC)
w5 <- waldtest(model_4,full,vcov = vcovHC)
w6 <- waldtest(simple,full,vcov = vcovHC)
tidy(w1)

F_comparison_table <- rbind(
  tidy(w1)[2, ] %>% mutate(Comparison = "simple vs model_2"),
  tidy(w2)[2, ] %>% mutate(Comparison = "model_2 vs model_3"),
  tidy(w3)[2, ] %>% mutate(Comparison = "model_2 vs model_4"),
  tidy(w4)[2, ] %>% mutate(Comparison = "model_3 vs full"),
  tidy(w5)[2, ] %>% mutate(Comparison = "model_4 vs full"),
  tidy(w6)[2, ] %>% mutate(Comparison = "simple vs full"),
deparse.level = 1) %>%
  select(Comparison, statistic, p.value, df, df.residual)

print(F_comparison_table)

f1 <- round(w1$F[2],3)
p1 <- w1$`Pr(>F)`[2]
f2 <- round(w2$F[2],3)
p2 <- w2$`Pr(>F)`[2]
f3 <- round(w3$F[2],3)
p3 <- w3$`Pr(>F)`[2]
f4 <- round(w4$F[2],3)
p4 <- w4$`Pr(>F)`[2]
f5 <- round(w5$F[2],3)
p5 <- w5$`Pr(>F)`[2]
f6 <- round(w6$F[2],3)
p6 <- w6$`Pr(>F)`[2]

get_stars <- function(p) {
  if (p < 0.01) return("***")
  if (p < 0.05) return("**")
  if (p < 0.1)  return("*")
  return("")
}

wald_row <- c("Wald F-test",
              " ",
              paste0(f1,get_stars(p1)), 
              paste0(f2,get_stars(p2)),
              paste0(f3,get_stars(p3)),
              #paste0(f4,get_stars(p4)),
              paste0(f5,get_stars(p5))#,
              #paste0(f6,get_stars(p6))
              )

```


```{r}
hf_dataset_resid <- hf_dataset %>%
  mutate(
    full_predictions = predict(full),
    full_residuals   = resid(full)
  )

plot_full <- hf_dataset_resid %>%
  ggplot(aes(x = full_predictions, y = full_residuals)) + 
  geom_point() + stat_smooth() +
  labs(title = "Residuals vs Fitted Values", y="Residuals", x= "Fitted Values") +
  theme_minimal()

plot_full

full_preds <- predict(full, new.data=hf_dataset, model_type="prediction",interval="confidence", level = 0.95)
hf_dataset_preds <- cbind(hf_dataset,full_preds)
hf_dataset_preds <- hf_dataset_preds %>%
  filter(hf_dataset_preds$training_stage != "Other")
view(hf_dataset_preds)

ggplot(hf_dataset_preds, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  geom_point() +
  geom_line(data=hf_dataset_preds, aes(y = fit)) +
  coord_cartesian(xlim = c(0, 2)) +
  geom_ribbon(data=hf_dataset_preds, aes(ymin = lwr, ymax = upr, y=fit), alpha = 0.2, color = NA)

ggplot(hf_dataset_preds, aes(x=log_size, y=fit, color=training_stage)) +
  #geom_point() +
  #geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  stat_smooth() +
  labs(title = "CI by Training Stage", y="Predicted Value", x= "log Parameters") + 
  theme_minimal()
```


```{r}
library(marginaleffects)

plot_predictions(full, condition= list("log_size","if_eval"=45.66, "training_stage"))
plot_slopes(full, variables = "training_stage", condition= list("log_size","if_eval"=45.66))
```


```{r}
# predicted values across size, if_eval held at median
pred_grid <- expand.grid(
  log_size = seq(min(hf_dataset$log_size),
                 max(hf_dataset$log_size),
                 length.out = 100),
  training_stage = levels(hf_dataset$training_stage),
  if_eval = 45.66 # median IFEval score
) %>%
  filter(training_stage != "Other")

pred_grid$pred_mmlu_pro <- predict(full, newdata = pred_grid)

ggplot(pred_grid, aes(x = log_size, y = pred_mmlu_pro, color = training_stage)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Log Parameter Count",
    y = "Predicted MMLU=Pro",
    color = "Training Stage"
  ) +
  theme_minimal()

pred_ci <- predict(full, newdata = pred_grid, interval = "confidence", level = 0.95)
pred_grid <- cbind(pred_grid, pred_ci)

dens <- density(hf_dataset$log_size, na.rm = TRUE, from = 0, to = 1)
dens_df <- data.frame(
  x = dens$x,
  y = 0.3 + dens$y / 50
)

  # x-axis labels 
  x_breaks <- c(seq(0, 0.8, 0.2), 0.95)
  x_labels <- c("0", "20", "40", "60", "80", "100%")

  # y-axis labels 
  y_breaks <- seq(0.2, 0.8, by = 0.1)
  y_labels <- c("20", "30", "40", "50", "60", "70", "80%")
stage_colors <- c(
  "Pretrained"     = "#E69F00",
  "Base-Merge"     = "#4A90D9",
  "Chat-Instruct"  = "#2C5F8A",
  "Fine-tuned"     = "#1A3A57"
)

ggplot(pred_grid, aes(x = log_size, y = fit, color = training_stage, fill = training_stage)) +
#  geom_line(data = dens_df, aes(x = x, y = y), linewidth = 0.7, color = "black") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 25.84, linetype = "dashed", color = "grey50") +
  annotate("text", x=-2, y=28, label="Mean MMLU-Pro Score", color="grey50") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  scale_x_continuous(
    breaks = c(seq(-2,1,1),2),
    labels = c("10M","100M","1B","10B","100B")
  ) +
  labs(
    x = "Parameter Count (log)",
    y = "Predicted MMLU-Pro",
    color = "Training Stage",
    fill = "Training Stage"
  ) +
  theme_minimal()
```

```{r}
# Repurposing Alex's code
# checking table we already used
library(data.table)
data.table(pred_grid)
########################

# creating predictions and CI
pred_data <- data.frame( 
  training_stage = levels(hf_dataset$training_stage), 
  log_size = rep(seq(from = min(hf_dataset$log_size, na.rm = T), to = max(hf_dataset$log_size, na.rm = T), length.out = 100) , 5),
  if_eval = 45.66 # median IFEval score
)

pred_data$fit     <- predict.lm(full, newdata = pred_data, interval = 'confidence')[ , 'fit']
pred_data$se.fit  <- predict.lm(full, newdata = pred_data, se.fit = T,)$se.fit
pred_data$low.ci  <- predict.lm(full, newdata = pred_data, interval = 'confidence', level = 0.95)[ , 'lwr']
pred_data$high.ci <- predict.lm(full, newdata = pred_data, interval = 'confidence', level = 0.95)[ , 'upr']

pred_data <- data.table(pred_data) %>%
  filter(training_stage != "Other")
pred_data

stage_colors <- c(
  "Pretrained"     = "#fdb515", # Orange: "#E69F00",
  "Base-Merge"     = "#4A90D9",
  "Chat-Instruct"  = "#2C5F8A",
  "Fine-tuned"     = "#003057" # Dark Blue replaced by Berkeley#1A3A57"
)

ggplot(pred_data, aes(x = log_size, y = fit, color = training_stage, fill = training_stage)) +
 # geom_line(data = dens_df, aes(x = x, y = y), linewidth = 0.7, color = "black") +
  geom_ribbon(aes(x=log_size,ymin = low.ci, ymax = high.ci), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 25.84, linetype = "dashed", color = "grey50") +
  annotate("text", x=-2, y=29, label="Mean MMLU-Pro Score", color="grey50") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  scale_x_continuous(
    breaks = c(seq(-2,1,1),2),
    labels = c("10M","100M","1B","10B","100B")
  ) +
  labs(
    title = "Predicted Performance By Training Stage",
    subtitle = "Full Model Prediction across LLM Size",
    caption = "Source: HuggingFace Open LLM Leaderboard Contents Dataset 2024",
    x = "Parameter Count (log)",
    y = "Predicted MMLU-Pro",
    color = "Training Stage",
    fill = "Training Stage"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"), plot.caption = element_text(hjust = 0))
ggplot(pred_grid, aes(x = log_size, y = fit, color = training_stage, fill = training_stage)) +
#  geom_line(data = dens_df, aes(x = x, y = y), linewidth = 0.7, color = "black") +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 25.84, linetype = "dashed", color = "grey50") +
  annotate("text", x=-2, y=29, label="Mean MMLU-Pro Score", color="grey50") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  scale_x_continuous(
    breaks = c(seq(-2,1,1),2),
    labels = c("10M","100M","1B","10B","100B")
  ) +
  labs(
    title = "Predicted Performance By Training Stage",
    subtitle = "Full Model Prediction across LLM Size",
    caption = "Source: HuggingFace Open LLM Leaderboard Contents Dataset 2024",
    x = "Parameter Count (log)",
    y = "Predicted MMLU-Pro",
    color = "Training Stage",
    fill = "Training Stage"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"), plot.caption = element_text(hjust = 0))

###################
#   # density line to be drawn inside plotting window
#   dens <- density(d$voteshare_republican_2016, na.rm = TRUE, from = 0, to = 1)
#   dens_df <- data.frame(
#     x = dens$x,
#     y = 0.3 + dens$y / 50
#   )
# 
#     # density line inside window
# geom_line(
#   data = dens_df,
#   aes(x = x, y = y),
#   linewidth = 0.7,
#   color = "black"
# ) +
#   
# ############  
# full_pred_type_plot_gg <- function(alternative = 3, title = "") {
# 
#   label_vector <- c("Latino Cue", "Black Cue", "Arab Cue")
#   
#   pal <- park_palette("Saguaro", 3)
#   pal_index <- c(3, 2, 1)
#   
#   alt_fill <- scales::alpha(pal[pal_index[alternative]], 0.9)
#   grey_fill <- scales::alpha("grey70", 0.5)
# 
#   # subset prediction data
#   pd_white <- pred_data |> filter(ethnic_cue == 0)
#   pd_alt   <- pred_data |> filter(ethnic_cue == alternative)
# 
#   # density line to be drawn inside plotting window
#   dens <- density(d$voteshare_republican_2016, na.rm = TRUE, from = 0, to = 1)
#   dens_df <- data.frame(
#     x = dens$x,
#     y = 0.3 + dens$y / 50
#   )
# 
#   # x-axis labels 
#   x_breaks <- c(seq(0, 0.8, 0.2), 0.95)
#   x_labels <- c("0", "20", "40", "60", "80", "100%")
# 
#   # y-axis labels 
#   y_breaks <- seq(0.2, 0.8, by = 0.1)
#   y_labels <- c("20", "30", "40", "50", "60", "70", "80%")
# 
#   # Build a combined data frame for legend mapping
#   ribbon_df <- bind_rows(
#     pd_alt   |> mutate(group = label_vector[alternative]),
#     pd_white |> mutate(group = "White Cue")
#   )
# 
#   line_df <- bind_rows(
#     pd_alt   |> mutate(group = label_vector[alternative]),
#     pd_white |> mutate(group = "White Cue")
#   )
# 
#   ggplot() +
#     # CI ribbons
#     geom_ribbon(
#       data = ribbon_df,
#       aes(
#         x = voteshare_republican_2016,
#         ymin = low.ci,
#         ymax = high.ci,
#         fill = group
#       ),
#       color = NA
#     ) +
#     # lines
#     geom_line(
#       data = line_df,
#       aes(x = voteshare_republican_2016, y = fit, linetype = group),
#       linewidth = 0.8,
#       color = "black"
#     ) +
#     # density line inside window
#     geom_line(
#       data = dens_df,
#       aes(x = x, y = y),
#       linewidth = 0.7,
#       color = "black"
#     ) +
#     # axes and labels
#     scale_x_continuous(
#       limits = c(0, 1),
#       breaks = x_breaks,
#       labels = x_labels
#     ) +
#     scale_y_continuous(
#       limits = c(0.3, 0.8),
#       breaks = y_breaks,
#       labels = y_labels
#     ) +
#     labs(
#       title = title,
#       x = "Trump Vote Share",
#       y = "Probability Respond"
#     ) +
#     scale_fill_manual(
#       values = c(
#         setNames(alt_fill, label_vector[alternative]),
#         "White Cue" = grey_fill
#       )
#     ) +
#     # dashed for alt, solid for white
#     scale_linetype_manual(
#       values = c(
#         setNames("dotted", label_vector[alternative]),
#         "White Cue" = "solid"
#       )
#     ) +
#     theme_classic(base_size = 14) +
#     theme(
#       legend.position = c(0.15, 0.15),
#       legend.justification = c(0, 0),
#       legend.title = element_blank(),
#       legend.background = element_blank(),
#       legend.key = element_rect(fill = NA, color = NA),
#       plot.title = element_text(hjust = 0.5)
#     )
# }
# 
# 
# library(tidyverse) 
# library(data.table)
# library(ggplot2) 
# library(patchwork) 
# library(nationalparkcolors)
# library(here) 
# 
# ## Load Analyis Package
# devtools::load_all()
# 
# ## Load data 
# d <- load_data(
#   filepath = here("data/processed/experiment_data.csv")
# )
# 
# election_returns <- fread(
#   file = here("data/processed/election_returns.csv"), 
#   colClasses = c("county_fips" = "character")
# )
# 
# setkeyv(x = election_returns, cols = "county_fips")
# d <- merge(d, election_returns, by = "county_fips", all.x = TRUE)
# 
# ## Estimate Model 
# model_ = d[ ,   lm(got_response ~ factor(ethnic_cue) * voteshare_republican_2016)]
# 
# ## Set up prediction data
# pred_data <- data.frame( 
#   ethnic_cue = factor(c(rep(0, 100), rep(1, 100), rep(2, 100), rep(3, 100))), 
#   voteshare_republican_2016 = rep(seq(from = min(d$voteshare_republican_2016, na.rm = T), to = max(d$voteshare_republican_2016, na.rm = T), length.out = 100) , 4)
# )
# 
# pred_data$fit     <- predict.lm(model_, newdata = pred_data, interval = 'confidence')[ , 'fit']
# pred_data$se.fit  <- predict.lm(model_, newdata = pred_data, se.fit = T,)$se.fit
# pred_data$low.ci  <- predict.lm(model_, newdata = pred_data, interval = 'confidence', level = 0.95)[ , 'lwr']
# pred_data$high.ci <- predict.lm(model_, newdata = pred_data, interval = 'confidence', level = 0.95)[ , 'upr']
# 
# pred_data <- data.table(pred_data)
# 
# ## Make Plots 
# latino_plot <- voteshare_hte_plot_gg(alternative = 1, title = "Latino Cue")
# black_plot  <- voteshare_hte_plot_gg(alternative = 2, title = "Black Cue")
# arab_plot   <- voteshare_hte_plot_gg(alternative = 3, title = "Arab/Muslim Cue")
# 
# (latino_plot + black_plot + arab_plot) + 
#   plot_layout(axis_titles = "collect")
# 
# ggsave(
#   filename = here("plots/hte_plot.png"), 
#   height = 4, 
#   width = 10, 
#   unit = "in", 
#   dpi = 300
#   )
```




```{r}
# Scatterplot with interaction lines
ggplot(df_no_other, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(title = "Scaling Laws: Performance by Model Size and Training Stage",
       x = "Log10(Parameters)",
       y = "MMLU-PRO Score",
       color = "Training Stage") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  theme_minimal()

# Fitted vs Residuals for Full Model
plot(full, which = 1, main = "Full Model: Residuals vs Fitted", pch=16, col="purple")
```
```{r}
# Scatterplot with interaction lines
ggplot(df_no_other, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  facet_wrap(~training_stage) +
  labs(title = "Scaling Laws: Performance by Model Size and Training Stage",
       x = "Log10(Parameters)",
       y = "MMLU-PRO Score",
       color = "Training Stage") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  theme_minimal()
```
```{r}
# Scatterplot with interaction lines
df_no_other_pred <- hf_dataset_resid %>%
  filter(hf_dataset$training_stage != "Other")
ggplot(df_no_other_pred, aes(x = log_size, y = mmlu_pro, color = training_stage)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  labs(title = "Scaling Laws: Performance by Model Size and Training Stage",
       x = "Log10(Parameters)",
       y = "MMLU-PRO Score",
       color = "Training Stage") +
  scale_color_manual(values = stage_colors) +
  scale_fill_manual(values = stage_colors) +
  theme_minimal()
# 
# library(sjPlot)
# library(sjmisc)
# library(ggplot2)
# plot_model(full,type="int") +
#   theme(theme_sjplot)
# 
# plot_model(full, type = "pred", terms = c("log_size","training_stage"))
```


```{r}
ggplot(hf_dataset_resid, aes(x = full_predictions, y = full_residuals)) + 
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) + 
  facet_wrap(~training_stage) + 
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted by Training Stage",
       x = "Fitted Values",
       y = "Residuals")

```
```{r}
hist(full_rse)
qqnorm(full_rse)
qqline(full_rse, lwd=2)
```
## Descriptive Statistics Table
```{r results='asis', echo=FALSE}
desc_stats <- hf_dataset %>% 
  group_by(training_stage) %>% 
  summarise(n = n(), 
            avg_size_B = mean(x_params_b),
            size_range = paste0(min(x_params_b), " - ", max(x_params_b)),
            avg_mmlu_pro = mean(mmlu_pro),
            avg_if_eval = mean(if_eval)
  )

desc_stats
suppressWarnings(
  stargazer(desc_stats,
            type = "text",
            column.labels = c("Training Stage", "Number of Models", "Avg Size", "Size Range", "Avg MMLU-Pro", "Avg IFEval", "Full Model"),
            summary = FALSE)
)
```


## Regression Table Output

```{r results='asis', echo=FALSE}
# Extract the robust standard errors column from each coeftest object
se_simple <- simple_rse[, 2]
se_m2 <- m2_rse[, 2]
se_m3 <- m3_rse[, 2]
se_m4 <- m4_rse[, 2]
se_full <- full_rse[, 2]

suppressWarnings(
  stargazer(simple, model_2, model_3, model_4, full, 
            type = "html", 
            title = "Model Performance Across Size, Training Stage, and Instruction Following",
            dep.var.labels = "MMLU-PRO Score",
            column.labels = c("Simple", "Main Effects", "+IFEval", "Interaction", "Full Model"),
            covariate.labels = c("Log(Parameters)", 
                                 "Chat-Instruct", 
                                 "Fine-tuned", 
                                 "Other", 
                                 "Pretrained", 
                                 "IFEval Score", 
                                 "Log(Params) × Chat-Instruct", 
                                 "Log(Params) × Fine-tuned", 
                                 "Log(Params) × Other", 
                                 "Log(Params) × Pretrained", 
                                 "Constant"),

            omit.stat = "f",
            add.lines = list(wald_row),
            notes = "Note: 'Base-Merge' is the reference category for Training Stage. Robust standard errors (HC3) reported in parentheses.",
            se = list(se_simple, se_m2, se_m3, se_m4, se_full), 
            header = FALSE)
)
